\chapter{Conclusion}

This thesis set out to investigate how socioeconomic factors like unemployment rate, educational attainment, GDP and defence spending influence active military personnel size in NATO member countries from 2015 to 2023. 
The analysis was conducted on a panel dataset, using a fixed-effects regression model.
Several findings were discovered for academic research and defence policy.

One of the most important and statistically significant findings was that a higher share of defence spending in a country's GDP is associated with a larger active armed forces size per capita.
This is particularly valuable as the share of defence spending in GDP is currently a relevant and widely discussed metric in the context of rising geopolitical tensions and shifting defence priorities.
In 2014, NATO Heads of State and Government committed to meeting a threshold of 2\% defence spending in GDP, and in 2025, all NATO countries are expected to meet or exceed this target \parencite{nato_defence_2025}. 
According to the findings, this increase could also result in larger active armed forces.

Secondary education attainment rates were significantly and negatively related to active military size per capita. 
This finding could mean that the military labour supply may decrease as populations become more educated. 
The results align with \textcite{hof_quality_2023}, who found that better-educated military recruits have higher intentions to quit basic training.
This negative relationship may reflect, for example, better civilian opportunities for more educated individuals or a shift toward more technology-oriented, rather than manpower-intensive armed forces.

Contrary to some prior studies, the unemployment rate did not exhibit statistically significant relationships with armed forces personnel. 
Military service is often regarded as a stable employment opportunity, especially in bad labour market conditions; however, this study found no evidence of labour market dynamics influencing military size. 
In contrast, GDP per capita, which can indicate better labour market conditions, as it showed a fairly strong negative correlation with unemployment rates, was found to be statistically significant and positively related to active military size. 
Other than labour market conditions, this finding may also reflect that wealthier countries can afford to maintain a larger military force, because they have more resources available for defence spending. 

The annual changes in GDP per capita and defence spending's share in GDP were found to be insignificant predictors of active military size. 
Short-term changes in GDP per capita may be too volatile to have direct effects on active military personnel, as defence planning could follow more long-term frameworks. 
The insignificance of annual changes in defence spending's proportion of GDP, contrasted by the significance of its overall level, may reflect that a sustained commitment to higher defence spending is necessary to grow military manpower, while short-term fluctuations may fund investments in other assets and therefore have limited immediate impact on military size.

For defence policy recommendations, the current study suggests keeping priority on sustained investments in the defence sector, as long-term commitments to a larger proportion of defence spending in GDP had a positive impact on military size. 
Additionally, GDP per capita had a significant positive relationship with armed forces per capita, highlighting the role of broader economic development in supporting military capacity. 
This suggests that defence planning should account for rising national income in supporting the maintenance of larger militaries.
It also recommends considering and addressing the reasons why educational attainment might negatively impact the military labour supply. 
New recruitment and retention strategies could be adapted to appeal to more educated populations.

A limitation of this study is that the independent variables themselves only explained about 20\% of the variation in active armed forces per capita with the fixed effects removed, while including fixed effects explained around 96\%.
This study aimed to investigate the relationships among the selected variables on active military size, which it successfully did; however, the variables themselves, without the fixed effects, do not predict a large portion of the variation in active armed forces per capita. 
For future studies looking to create a better prediction model, more variables, and for example, the effect of defence planning policy or the technological evolution of militaries should be taken into account, as personnel size may not depend only on socioeconomic factors.
Another limitation is the sample size. According to \textcite{harrell_multivariable_2015}, in order to estimate a reliable multivariate regression, the data should have at least 10-20 observations per estimated parameter, which this model did. 
However, to increase the statistical power of a fixed-effects regression, future research could collect more observations per country or include additional countries for more generalizable results.
An additional future research opportunity might be investigating the reasons why exactly educational attainment has a negative correlation to active military size, as this study does not reveal the causal effects.