\chapter{Methodology}

\section{Data preparation}

The data preparation process involved several steps to create a clean and structured 
dataset that could be used for analysis. The study combines military and socioeconomic 
data for NATO countries across multiple years. Data was collected from various sources,
transformed into a consistent format and merged into a single dataset.

First, the military personnel and defence 
spending data were acquired from different issues of The Military Balance by the 
\textcite{noauthor_military_balance_nodate} 
as each issue contained data for a specific year. Data for active armed forces numbers, 
defence spending per capita and defence spending's share of GDP were
selected for NATO member countries from each table.
In order to ensure that the data is comparable, the effect of inflation was removed 
from defence spending per capita data by adjusting the values to 2015 USD using the Consumer Price 
Index (CPI) of different years. The CPI data was sourced from \textcite{federal_reserve_bank_of_minneapolis_consumer_nodate}. 
The tables from different years were then merged into a single long-format table, where
each row represented observations for a specific country in a specific year, as this format is 
suitable for panel data analysis.

Next, the unemployment rate, GDP per capita in constant 2015 USD, population and 
secondary educational attainment data were collected from the open data provided by the
\textcite{noauthor_world_bank_nodate}. Data was again filtered to include 
only NATO countries from 2015 to 2023. 
The educational attainment 
data had missing values that were handled using linear interpolation to estimate missing values 
based on neighbouring data, and backward/forward filling for data points at the start or end 
of the series, which could not be interpolated. The tables were then also transformed into a long format, 
so that they could be 
merged with the military personnel and defence spending data.

The military and economic data were then merged together into a complete dataset. 
The active armed forces per capita column was created by dividing the active armed forces by population numbers.
Additional columns for annual changes in GDP per capita, defence spending per capita  
and defence spending's share of GDP were calculated by finding their percentage changes from 
the previous year's value. Table~\ref{tab:variables} summarizes the variables of the prepared dataset.

\renewcommand{\arraystretch}{1.3}

\begin{table}[ht]
\small
\caption{Variable Descriptions}
\centering
\resizebox{\textwidth}{!}{%
\begin{tabularx}{\textwidth}{>{\hsize=1\hsize\raggedright\arraybackslash}X 
                            >{\hsize=1.75\hsize\raggedright\arraybackslash}X 
                            >{\raggedright\arraybackslash}X 
                            >{\hsize=0.6\hsize\raggedright\arraybackslash}X 
                            >{\hsize=0.6\hsize\raggedright\arraybackslash}X}
\toprule
\textbf{Variable} & \textbf{Description} & \textbf{Source} & \textbf{Unit} & \textbf{Transformation} \\
\midrule
Active armed forces per capita & Number of active military personnel per capita & Military Balance, World Bank & per capita & log() \\
Unemployment rate & National unemployment rate & World Bank & \% (0-100) & none \\
Secondary education rate & Proportion of population (25+) with secondary education & World Bank & \% (0-100) & none \\
GDP per capita & Gross Domestic Product per capita & World Bank & 2015 USD & log() \\
Defence spending per capita & A country's defence expenditure divided by population & Military Balance & 2015 USD & log() \\
Defence spending \% of GDP & Defence spending's share of GDP & Military Balance & \% & none \\
GDP per capita \% change & Annual percentage change in GDP per capita & World Bank (own calculation) & \% & none \\
Defence spending per capita \% change & Annual percentage change in defence spending per capita & Military Balance (own calculation) & \% & none \\
Defence spending \% GDP \% change & Annual percentage change in defence spending's share of GDP & Military Balance (own calculation) & \% & none \\
\bottomrule
\end{tabularx}
}
\label{tab:variables}
\end{table}

Finally, a logarithmic transformation was applied to variables active armed forces per capita, GDP per capita 
and defence spending per capita. These variables can have skewed distributions due to disproportionate influence 
from larger countries. 
Logging can mitigate this skewness, stabilise variance and allow for interpretation of the regression 
coefficients in terms of elasticities. Other variables were not logarithmically transformed as they were 
expressed as percentages; therefore, logging would make interpretation difficult and distort the scales.

\section{Data understanding}

Table~\ref{tab:descriptive_stats} reports the descriptive statistics for the variables used.
The logged dependent variable, armed forces per capita, showed relatively low variation with a standard 
deviation of 0.44 and a range of values between -6.55 and -4.30, which could limit the explanatory 
power of the regression model, however, the regression model might still provide insights into 
smaller changes in active military size per capita.

\begin{table}[htbp]
\caption{Descriptive statistics of variables}
\renewcommand{\arraystretch}{1.2}
\begin{tabularx}{\textwidth}{l*{9}{>{\centering\arraybackslash}X}}
\toprule
 & \rotatebox{90}{\parbox{2.5cm}{Armed Forces per cap.}} 
 & \rotatebox{90}{\parbox{2.5cm}{GDP per cap.}} 
 & \rotatebox{90}{\parbox{2.5cm}{Def. spend. per cap.}} 
 & \rotatebox{90}{\parbox{2.5cm}{Unemploy-\\ment rate}} 
 & \rotatebox{90}{\parbox{2.5cm}{Secondary education rate}} 
 & \rotatebox{90}{\parbox{2.5cm}{Def. spend. \% GDP}} 
 & \rotatebox{90}{\parbox{2.5cm}{GDP per cap. \% change}} 
 & \rotatebox{90}{\parbox{2.5cm}{Def. spend. per cap. \% change}} 
 & \rotatebox{90}{\parbox{2.5cm}{Def. spend. \% GDP \% change}} \\
\midrule
\textbf{mean} & -5.71 & 7.86 & 74.50 & 10.03 & 5.71 & 1.61 & 2.18 & 4.82 & 4.57 \\
\textbf{std} & 0.44 & 4.58 & 14.64 & 0.76 & 0.81 & 0.66 & 3.67 & 18.73 & 16.39 \\
\textbf{min} & -6.55 & 2.02 & 33.60 & 8.28 & 3.44 & 0.35 & -15.21 & -41.71 & -41.03 \\
\textbf{25\%} & -5.96 & 4.83 & 69.07 & 9.52 & 5.26 & 1.14 & 0.76 & -5.61 & -3.59 \\
\textbf{50\%} & -5.79 & 6.54 & 77.78 & 9.93 & 5.81 & 1.44 & 2.27 & 1.89 & 1.48 \\
\textbf{75\%} & -5.46 & 9.41 & 86.00 & 10.70 & 6.29 & 1.98 & 4.37 & 10.16 & 10.67 \\
\textbf{max} & -4.30 & 26.40 & 95.29 & 11.59 & 7.60 & 3.82 & 13.65 & 100.83 & 109.23 \\
\bottomrule
\end{tabularx}
\label{tab:descriptive_stats}
\end{table}

Defence spending per capita, its annual change and the annual change in defence spending's 
share of GDP however showed higher volatility, with standard deviations of 14.64, 18.73, and 
16.39 respectively. These higher standard deviations could suggest heteroskedasticity. In addition 
to that, in panel data settings, standard errors are often correlated within entities (in this case countries), 
which means assuming independent standard errors could inflate t-statistics and potentially 
overstate statistical significance in regressions. To account for this, a fixed-effects 
regression could be estimated using standard errors clustered at the country level.

The use of imputed educational attainment values was also investigated. There were a total 
of 23 interpolated or filled values across a total of 285 observations. No country had
more than 2 imputed values in the timeframe. Given the small scale of imputed values, it was 
unlikely that a strong bias would be introduced. Nevertheless, a sensitivity analysis should 
be conducted to analyse the effect of imputation more in depth.

Outliers were identified using both Z-score and 
Interquartile Range (IQR) methods, which are commonly used in literature for outlier 
detection. The Z-score method calculates the distance of a data point from 
the mean in terms of standard deviations, using the formula:
\begin{equation*}
    Z = \frac{X - \mu}{\sigma}
\end{equation*}
where X is the data point, $\mu$ is the sample mean and $\sigma$ is the sample standard deviation.
A common threshold of $\pm 3$ was used to identify values in approximately the outer 0.3\% of normally 
distributed data.
The IQR method is non-parametric and less dependent on distributional assumptions. The outliers 
were defined as values, which fall below $Q1 - 1.5 \cdot IQR$ or above $Q3 + 1.5 \cdot IQR$, where 
$Q1$ and $Q3$ are the first and third quartiles, reported in Table~\ref{tab:descriptive_stats} as 
25\% and 75\%, and $IQR = Q3 - Q1$. This approach helps identify skewed or non-normally distributed outliers.
Comparing outliers to the complete dataset revealed no unrealistic 
values. Since the outliers appeared to be the result of natural variability, not data 
quality issues, they were not excluded from the dataset. \parencite{barnett_outliers_1994}

\section{Correlation analysis}

This section describes the analytical approach used to investigate the relationships between 
socioeconomic variables and active military personnel size.
It describes how correlations and multicollinearity of the data were evaluated, 
how a sensitivity analysis for imputed values and a robustness check for multicollinear 
variables were conducted. Finally, it is explained how the regression model was estimated.

The correlation and multicollinearity analysis were performed on demeaned data, consistent 
with the fixed-effects regression. The fixed-effects model relies on within-country variation 
over time by subtracting country-specific means, so the data were transformed similarly. For 
each observation, the mean of the variable across all years of the corresponding country was 
subtracted from the original value. This demeaning process makes the data comparable across 
countries as the variables only reflect deviations from each country's average. 

The correlation analysis involved calculating the Pearson correlation coefficient for each 
of the variables, resulting in a correlation matrix, which can be used to assess the 
direction and strength of bivariate relationships. 
The Pearson coefficient, denoted by $r$, is 
defined as:
\begin{equation*}
    r_{xy} = \frac{\sum(x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum(x_i - \bar{x})^2 \sum(y_i - \bar{y})^2}}
\end{equation*}
where $x_i$ and $y_i$ are observations of variables $X$ and $Y$, and $\bar{x}$, $\bar{y}$ are their 
sample means. The coefficient values range from -1 to 1, where absolute values close to 1 indicate 
strong relationships and values near 0 suggest weak or no correlations \parencite{cohen_bivariate_2003}. 

Variance Inflation Factors (VIF) were calculated for the independent variables to 
assess the potential presence of multicollinearity. VIFs indicate how much the variance
of a regression coefficient is inflated due to relationships with other predictors. Multicollinearity 
can make estimates unstable and cause interpretability problems. The VIF for 
a variable $X_i$ is calculated as:
\begin{equation*}
    VIF_i = \frac{1}{1-R^2_i}
\end{equation*}
where $R^2_i$ is the coefficient of determination obrained by regressing $X_i$ on all other independent
variables. It is typically considered, that VIF values from 1 to 5 are
moderate and mostly acceptable, while values over 5, or more conservatively 10, 
can be potentially problematic \parencite{obrien_caution_2007}.

\section{Modelling}

Studies such as \textcite{asch_cash_2010}, \textcite{backstrom_are_2019} and 
\textcite{balcaen_unemployment_2025} have demonstrated the suitability of a fixed-effects 
regression model for analysing panel data in the context of military recruitment and retention, 
which informed the methodolical choice of this thesis.
The following equation was used to estimate the base model:
\begin{align*}
log(ArmedForces_{it}) &= \beta_1 \cdot Unemployment_{it} 
+ \beta_2 \cdot Education_{it} \\
&\quad + \beta_3 \cdot log(GDPPerCap_{it}) 
+ \beta_4 \cdot log(DefSpendingPerCap_{it}) \\
&\quad + \beta_5 \cdot DefSpending\%GDP_{it} 
+ \beta_6 \cdot \Delta GDPPerCap_{it} \\
&\quad + \beta_7 \cdot \Delta DefSpendingPerCap_{it} 
+ \beta_8 \cdot \Delta DefSpending\%GDP_{it} \\
&\quad + \alpha_i + \gamma_t + \varepsilon_{it}
\end{align*}
where $i$ denotes the country, $t$ the time, $\Delta$ the annual change,
$a_i$ is the country-specific fixed effect and $\gamma_t$ is the year fixed effect.

The regression analysis used a fixed-effects panel regression model to control for time-invariant 
heterogeneity across countries. This meant that it was not 
necessary to use the demeaned data anymore, 
because the model already took differences between entities into account.
The regression model used country fixed effects to isolate the within-country variation over time, 
enabling more accurate 
estimation of the conditional effects of economic factors on military personnel levels.
In other words, by introducing country fixed effects, the model was estimated on changes within 
countries, rather than between countries \parencite{backstrom_are_2019}.

Time fixed effects were also included in the model to control for unobserved heterogeneity
over time. Given global shocks, such as the COVID-19 pandemic and Russia's invasion of 
Ukraine, which could have affected all countries similarly, year fixed effects were 
necessary to avoid omitted variable bias and eliminate aggregate trends in the data \parencite{backstrom_are_2019}.
For comparison, a model with only country fixed effects and no time fixed effects
is reported in Table~\ref{tab:oneway_summary} and Table~\ref{tab:oneway_model} in 
the \textit{Appendix}.
It can be used to help assess the impact of including time fixed effects.

The final fixed-effects regression model used clustered country-level standard errors to 
account for potential within-country correlation of error terms and heteroskedasticity across 
countries. In panel data, observations from the same entity are often not independent, which 
violates classical OLS assumptions. Clustering allows for arbitrary correlation within 
countries, ensuring a more robust and conservative model. While common trends may 
affect all countries in a given year, clustering at the time-level was not used, as 
the data has a limited number of time periods, which could lead to unstable clustering 
or an overly conservative model.

A robustness check and a sensitivity analysis were conducted to further specify the model.
The robustness check included estimating two models, one with full variables and one with 
reduced variables, based on the results of the correlation analysis. The models were 
compared to assess the impact of removing multicollinear variables.
The sensitivty analysis was used determine whether to include interpolated and filled educational 
attainment data in the regression model. Three regression models were created: Model A with 
complete data, including observations with interpolated educational attainment values, Model B 
that excluded rows with interpolated and filled values, and Model C with complete data and an additional 
dummy variable indicating whether the education data was interpolated or not. The models 
were again compared to choose the most appropriate specification.